Analysis

### 2023-07-05 analysis/analysis.Rmd
Revise section and chunk titles

## Loading the necessary libraries

@@

```{r message=FALSE}
library(analogue)
library(coda)
library(doParallel)
library(fmcmc)
library(ggdist)
library(ggspatial)
library(grid)
library(here)
library(MCMCvis)
library(neotoma2)
library(nimble)
library(rcarbon)
library(RcppRoll)
library(reshape2)
library(sf)
library(tidyverse)
library(vegan)
```

@@

```{r message=FALSE}
if(!require('devtools')) install.packages('devtools')
library(devtools)
install_github('ISAAKiel/aoristAAR')
require(aoristAAR)
```

```{r}
source(here("code/map_helpers.R"))
```

## Data collection

@@

### Openness index

@@

```{r}
pollen_locations <- read.csv(here("data","raw_data", "pollen_locations.csv"))
pollen_locations
```

@@

```{r}
pollen_locations.sf <- st_as_sf(pollen_locations, 
                                coords = c("lng", "lat"), crs = 4326)

@@

We download the data from neotoma:

```{r message=FALSE, warning=FALSE}
pollen_data <- get_downloads(pollen_locations$neotoma_id, all_data=T)
```

Then we load a few help scripts which can be found under "code/neotoma_helpers.R".

```{r}
source(here("code", "neotoma_helpers.R"))
```

Now we filter the data sets so that we only retain the pollen data that record "NISP".

```{r}
pollen_data.samples <- samples(pollen_data)

pollen_data.samples.filtered <- pollen_data.samples %>% subset(element == "pollen" & units == "NISP")
```

Then we convert the NISP into percentages:

```{r}
pollen_data_comp <- transform_to_percent(pollen_data.samples.filtered)
```

@@

### ^14^C Summenkalibration

The data set for the 14C sum calibration is composed of the above-mentioned sources. This data consists of 1135 dates from 246 sites.

```{r}
c14 <- read.csv(here("data","raw_data", "14c_swiss_plateau.csv"))
```

@@

```{r}
c14.sf <- c14 |>
  distinct(site, lat, lng) |>
  st_as_sf(coords = c("lng", "lat"),  crs = 4326)

@@

```{r}
c14.calibrated <- calibrate(x = c14$bp, errors = c14$std, verbose = F)
```

@@

```{r}
sumcal <- spd(c14.calibrated, timeRange = 1950-timeframe, bins = c14$site, runm = sw, verbose = F)
```

@@

```{r}
plot(1950-sumcal$grid$calBP, sumcal$grid$PrDens,type="l")
```

@@

```{r}
res.sumcal <- sumcal$grid[sumcal$grid$calBP %in%
seq(from=1000,
    to=10000, by=50),]
```

### Aoristic Site Count

@@

```{r}
sites <- read.csv(here("data","raw_data", "site_data_fuzzy.csv"))
```

@@

```{r}
sites.sf <- st_as_sf(sites,
                   coords = c("LV03_Rechtswert_fuzzy", "LV03_Hochwert_fuzzy"),
                   crs = 21781)

@@

### Dendro Dated Settlements

@@

```{r}
lss <- read.csv(here("data","raw_data", "dendro_years.csv"))
```

@@

```{r}
plot(res.lss$date, res.lss$moving_mean,type="l")
```

@@

```{r}
res.lss <- res.lss[res.lss$date %in% seq(from=timeframe[1],
                                         to=timeframe[2], by=50),
        ]
res.lss$age <- 1950-res.lss$date
```

### Combining the proxies

@@

```{r}
ggplot() +
  geom_line(data = res.open_final, aes(x = 1950-age, y = scale(mean)), col="darkgreen", alpha = 0.75) +
  geom_line(data = res.sumcal, aes(x = 1950-calBP, y = scale(PrDens)), col="red", alpha = 0.75) +
  geom_line(data = res.aorist, aes(x = date, y = scale(sum)), col="gray", alpha = 0.75) + 
  geom_line(data = res.lss, aes(x = date, y = scale(moving_mean)), col="brown", alpha = 0.75) +
  scale_colour_manual(values = c("darkgreen", "red", "gray", "brown"),
                      labels = c("Openness Indicator", "Sum Calibration", "Aoristic Sum", "Dendro Dated Settlements")
                      )
```

@@

```{r}
all_proxies <- merge(res.sumcal, res.open_final, by.x = "calBP", by.y = "age") %>%
  merge(res.aorist, by.x = "calBP", by.y = "age") %>%
  merge(res.lss, by.x = "calBP", by.y = "age") %>%
  select(calBP, PrDens, mean, sum, moving_mean) %>%
  rename(age = calBP, sumcal = PrDens, openness = mean, aoristic_sum = sum, dendro = moving_mean) %>%
  dplyr::filter(age < 8000 & age > 3000)

@@

```{r}
write.csv(all_proxies, here("data","preprocessed_data", "all_proxies.csv"))
```

## Transforming the Data

@@

```{r}
all_proxies <- read.csv(here("data","preprocessed_data", "all_proxies.csv"),
                        row.names = 1)
```

@@

## The Model

@@

```{r}
model_code <- nimbleCode( {
  # ---- Process Model ----

@@

```{r}
model_constants <- list(
  # an estimation for the population density at the end of the modeled period
  nEnd = 5, 
  # the number of time slices
  nYears = redux_nrow,
  # the area of the swiss plateau for the transformation
  # of population density to Total Population
  AreaSwissPlateau = 12649
  )
```

@@

# Calculation of an exponential increase of the factor 10
r <- log(1 - (10^(1/model_constants$nYears) - 1))
model_inits <- list(
  lambda = rep(r,redux_nrow),
  PopDens = rep(model_constants$nEnd,redux_nrow),
  z = rep(50,redux_nrow)
)
```

@@

```{r}
batches <- 100000 # run in batches of this length
thinning <- 10 # save only each ... of the chains
```

@@

```{r}
model_monitors <- c("PopDens", "beta")
```

@@

@@

```{r}
start_time <- Sys.time()

@@

```{r}
# Initialize with non-convergence
converged <- F

@@

```{r}
do_final_run = F
```

>>>>

## Dependencies

@@

```{r dependencies, message=FALSE}
library(analogue)
library(coda)
library(doParallel)
library(fmcmc)
library(ggdist)
library(ggspatial)
library(grid)
library(here)
library(MCMCvis)
library(neotoma2)
library(nimble)
library(rcarbon)
library(RcppRoll)
library(reshape2)
library(sf)
library(tidyverse)
library(vegan)
```

@@

```{r dev-dependencies, message=FALSE}
if(!require('devtools')) install.packages('devtools')
library(devtools)
install_github('ISAAKiel/aoristAAR')
require(aoristAAR)
```

```{r source-dependencies}
source(here("code", "map_helpers.R"))
source(here("code", "neotoma_helpers.R"))
```

## Proxy data

@@

### Pollen (openness index)

@@

```{r data-pollen-locations}
(pollen_locations <- read.csv(here("data","raw_data", "pollen_locations.csv")))
```

@@

```{r map-pollen-locations}
pollen_locations.sf <- st_as_sf(pollen_locations, 
                                coords = c("lng", "lat"), crs = 4326)

@@

We download data for these locations using the neotoma2 package and some helper functions defined in `code/neotoma_helpers.R`.
We retain only records that include the NISP, which we convert into a percentage.

```{r data-pollen, message=FALSE, warning=FALSE}
pollen_data <- get_downloads(pollen_locations$neotoma_id, all_data=T)
pollen_data.samples <- samples(pollen_data)
pollen_data.samples.filtered <- pollen_data.samples %>% subset(element == "pollen" & units == "NISP")
pollen_data_comp <- transform_to_percent(pollen_data.samples.filtered)
```

@@

### Summed radiocarbon

The data set for the ^14^C sum calibration is composed of the above-mentioned sources. This data consists of 1135 dates from 246 sites.

```{r read-c14}
c14 <- read.csv(here("data","raw_data", "14c_swiss_plateau.csv"))
```

@@

```{r plot-c14}
c14.sf <- c14 |>
  distinct(site, lat, lng) |>
  st_as_sf(coords = c("lng", "lat"),  crs = 4326)

@@

```{r calibrate-c14}
c14.calibrated <- calibrate(x = c14$bp, errors = c14$std, verbose = F)
```

@@

```{r sum-c14}
sumcal <- spd(c14.calibrated, timeRange = 1950-timeframe, bins = c14$site, runm = sw, verbose = F)
```

@@

```{r plot-c14-sum}
plot(1950-sumcal$grid$calBP, sumcal$grid$PrDens,type="l")
```

@@

```{r resample-c14-sum}
res.sumcal <- sumcal$grid[sumcal$grid$calBP %in%
seq(from=1000,
    to=10000, by=50),]
```

### All settlements

@@

```{r read-sites}
sites <- read.csv(here("data","raw_data", "site_data_fuzzy.csv"))
```

@@

```{r plot-sites}
sites.sf <- st_as_sf(sites,
                   coords = c("LV03_Rechtswert_fuzzy", "LV03_Hochwert_fuzzy"),
                   crs = 21781)

@@

### Lakeshore settlements

@@

```{r read-lss}
lss <- read.csv(here("data","raw_data", "dendro_years.csv"))
```

@@

```{r plot-lss}
plot(res.lss$date, res.lss$moving_mean,type="l")
```

@@

```{r resample-lss}
res.lss <- res.lss[res.lss$date %in% seq(from=timeframe[1],
                                         to=timeframe[2], by=50),
        ]
res.lss$age <- 1950-res.lss$date
```

## Combined proxies

@@

```{r plot-proxies}
ggplot() +
  geom_line(data = res.open_final, aes(x = 1950-age, y = scale(mean)), col="darkgreen", alpha = 0.75) +
  geom_line(data = res.sumcal, aes(x = 1950-calBP, y = scale(PrDens)), col="red", alpha = 0.75) +
  geom_line(data = res.aorist, aes(x = date, y = scale(sum)), col="gray", alpha = 0.75) + 
  geom_line(data = res.lss, aes(x = date, y = scale(moving_mean)), col="brown", alpha = 0.75) +
  scale_colour_manual(values = c("darkgreen", "red", "gray", "brown"),
                      labels = c("Openness Indicator", "Sum Calibration", "Aoristic Sum", "Dendro Dated Settlements")
                      )
```

@@

```{r combine-proxies}
all_proxies <- merge(res.sumcal, res.open_final, by.x = "calBP", by.y = "age") %>%
  merge(res.aorist, by.x = "calBP", by.y = "age") %>%
  merge(res.lss, by.x = "calBP", by.y = "age") %>%
  select(calBP, PrDens, mean, sum, moving_mean) %>%
  rename(age = calBP, sumcal = PrDens, openness = mean, aoristic_sum = sum, dendro = moving_mean) %>%
  dplyr::filter(age < 8000 & age > 3000)

@@

```{r cache-proxies}
write.csv(all_proxies, here("data","preprocessed_data", "all_proxies.csv"))
```

### Transforming the Data

@@

```{r read-cached-proxies}
all_proxies <- read.csv(here("data","preprocessed_data", "all_proxies.csv"),
                        row.names = 1)
```

@@

## Model definition

@@

```{r define-model}
model_code <- nimbleCode( {
  # ---- Process Model ----

@@

```{r define-model-constants}
model_constants <- list(
  # an estimation for the population density at the end of the modeled period
  nEnd = 5, 
  # the number of time slices
  nYears = redux_nrow,
  # the area of the swiss plateau for the transformation
  # of population density to Total Population
  AreaSwissPlateau = 12649
  )
```

@@

```{r define-model-inits}
# Calculation of an exponential increase of the factor 10
r <- log(1 - (10^(1/model_constants$nYears) - 1))
model_inits <- list(
  lambda = rep(r,redux_nrow),
  PopDens = rep(model_constants$nEnd,redux_nrow),
  z = rep(50,redux_nrow)
)
```

@@

```{r define-run-params}
batches <- 100000 # run in batches of this length
thinning <- 10 # save only each ... of the chains
```

@@

```{r define-run-monitors}
model_monitors <- c("PopDens", "beta")
```

@@

```{r setup-parallel}

@@

```{r run-model-inital}
start_time <- Sys.time()

@@

```{r run-model-convergence}
# Initialize with non-convergence
converged <- F

@@

```{r run-model-final}
do_final_run = F
```
